Drought regimens predict life history strategies in \emph{Heliophila}

### 2019-06-08 manuscript/manuscript.Rmd
proofing

author: 
  - name          : "J. Grey Monroe"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "307 University Ave, Fort Collins, CO 80521"
    email         : "monroejg@colostate.edu"
  - name          : "Brian Gill"
    affiliation   : "3"
  - name          : "Kathryn G. Turner"
    affiliation   : "4"
  - name          : "John K. McKay"
    affiliation   : "2"

@@

keywords          : "drought adaptation, herbaria records, Heliophila, life history evolution, phylogeography, remote sensing "

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf

@@

Understanding the causes and consequences of life history variation is a longstanding goal of ecology and evolutionary biology  [@cole1954population]. In  plants, life histories are especially diverse, with some species completing their life cycle in a number of weeks to others that live for thousands of years [@brown1996oldlist]. Along this continuum in angiosperms an important division exists distinguishing annuals which complete their seed to seed life cycle within a single calendar year from perennials which can persist over multiple years. Annual plants do not need to survive through the full range of seasonal environmental variation and spend at least some portion of the year as a seed where they are relatively protected from environmental stress. In contrast, perennial plants can continue vegetative growth over multiple years and must survive conditions experienced during all seasons but can also benefit from competitive advantages and, if iteroparous, multiple bouts of reproduction. These represent fundamentally different life history strategies and predicting their occurrence is important for community, ecosystem, and agricultural ecology. However, the environmental factors that explain their evolution and distributions remain empirically unresolved [@friedman2015all].

Classical theory predicts shorter life spans in environments where adult mortality is high [@charnov1973life;@stearns1992evolution;@franco1996life]. Because lack of water is perhaps the greatest threat to survival during vegetative or reproductive growth in plants, this theory has been extended to the hypothesis that annuality is adaptive when it allows plants to escape drought [@schaffer1975selection]. Indeed, adaptation to drought,  defined as episodes of increased aridity causing plant sress [@passioura1996drought], is often invoked as an explanation for the success of annual species. And while a few cases are cited where annuality appears to be more common in environments with greater aridity [@stebbins1952aridity; @morishima1984differentiation], this hypothesis has yet to be supported while controlling for the effect of common ancestry (phylogeny) on life habit. In one previous study where this question was addressed phylogenetically,  [@evans2005climate] annuals were not found to be associated with environments that experience more drought. This could be explained by the relatively small number of species studied and the reliance on a limited number of weather stations to characterize environments, highlighting the need to develop more scalable methods to study the geographic distributions of traits such as life history. Thus, in this study we leverage thousands of herbaria specimens among dozens of species and high-resolution remote sensing to study the distributions and environmental factors potentially driving the evolution and distribution of life history.

It is also critical to consider another dimension of drought adaptation: the expectation that annuality is most adaptive when droughts are not only frequent but also predictable. That is, when the frequency of drought is particularly high during certain seasons. Such predictability is important for selection to favor and escape strategy during those seasons which are particularly drought prone. While there has been at least one example of annuality associated with environments qualitatively classified as "predictable" in a general sense [@datson2008climate], the seasonal predictability of drought experienced by annuals has yet to be rigorously studied. As such, further empirical work is needed to support the model of annuality as a mechanism of drought adaptation via escape from drought prone seasons. Here we study herbarium collection dates to ask whether annuals indeed exhibit evidence of an escape strategy from seasons with elevated drought frequency.

@@

Here we combine a long-term global dataset of satellite detected drought events with metadata from natural history collections to test these classic hypotheses within the African endemic mustard genus, _Heliophila_ L. (Brassicaceae). If annuality is an adaptive strategy allowing plants to escape drought prone seasons, then drought frequency should predict the distribution of life history strategies across landscapes, and annual species should be more commonly associated with drought prone regions than perennial species. Additionally, if perenniality offers competitive advantage in the absence of drought, associations between life history and drought frequency should be significant when phylogenies include transitions from annual to perennial life history strategy. Finally, if annual species have adapted to escape predictably drought prone seasons, observations of growing annual species (i.e. occurring in forms other than seed) should be rare during seasons when drought frequency is highest. Phylogenetic relatedness can influence tests of associations between species’ traits and their environments by confounding common environments caused by selection from common environments caused my ancestry. [@barrett1996comparative;@felsenstein1985phylogenies]. Therefore, we assessed the relationship between life history distribution and drought frequency while controlling for phylogeny.

@@

### _Heliophila_ occurrence records
```{r raw_GBIF}
GBIF<-read.csv("../data/0006178-160311141623029/occurrence.csv")
```
To characterize the distributions of annual (studied here, n = 21) and perennial (studied here, n = 21) _Heliophila_ species, all (`r nrow(GBIF)`) records for the genus _Heliophila_ were downloaded from the Global Biodiversity Information Facility (gbif.org) on July 21, 2018 [@gbifdownload]. Herbaria records such as these provide a rich data sources to characterize the geographical distributions of species [@thiers2016index; @willis2017old; @lang2018using]. And as they become digitized [@soltis2017digitization], herbaria collections have been used to study relationships between trait distributions, geography, and climate [@davis2015herbarium;@wolf2016altitudinal;@stropp2016mapping;@vaclavik2017effects].

@@

```{r Droughtdata}
Winter<-raster("../data/drought/Summer_drought_freq")
Spring<-raster("../data/drought/Fall_drought_freq")
Summer<-raster("../data/drought/Winter_drought_freq")
Fall<-raster("../data/drought/Spring_drought_freq")

# Winter<-raster("~/VHI/droughtfreq/summer_mild")
# Spring<-raster("~/VHI/droughtfreq/fall_mild")
# Summer<-raster("~/VHI/droughtfreq/winter_mild")
# Fall<-raster("~/VHI/droughtfreq/spring_mild")
```

@@

$$VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})$$
By combining VCI and TCI, the VHI distinguishes drought from other forms of vegetative stress [@kogan1995application]. The use of the VHI to detect drought has been validated globally and across ecosystem types [@aghakouchak2015remote], including in southern Africa, the focal region of this study (e.g. Figure \@ref(fig:mapsdroughtexamples)). To date, the VHI has most often been applied for evaluating drought risk for agricultural research [e.g., @rojas2011assessing; @kogan2016modelling]. But it also presents a tool to study seasonal patterns in the frequency of drought across environments and to test hypotheses about the effect of drought on ecological and evolutionary processes [@kerr2003space]. As such, the VHI  has been applied recently to study drought related ecology of natural species and proven useful for predicting intraspecific variation in drought tolerance traits and genes [@Mojica2016;@dittberner2018natural;@monroe2018drought]. Here, we accessed VHI data at $16km^2$ resolution from 1981 to 2015 (https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php) to characterize the seasonal drought frequencies experienced by annual and perennial _Heliophila_ species across their native range of southern Africa.

## Analyses
### Drought frequency calculations
To characterize drought regimens across the distributions of annual and perennial species of _Heliophila_, we calculated drought during different seasons at the location of observations for _Heliophila_ records using the VHI. Specifically, we created maps of the frequencies of observing drought conditions between years (VHI<40, NOAA) during the winter (quarter surrounding winter solstice), spring (quarter surrounding spring equinox), summer (quarter surrounding summer solstice) and fall (quarter surrounding fall equinox) from 1981 to 2015 across the range of _Heliophila_. From these maps, the drought frequency (the number of times drought is observed divided by the total number of years, 34) during the winter, spring, summer, and fall were extracted for the locations of all GBIF records.

@@

GBIF <- GBIF %>%
  mutate(Winter=raster::extract(Winter, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(Spring=raster::extract(Spring, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(Summer=raster::extract(Summer, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(Fall=raster::extract(Fall, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(ymd_lat_lon_sp = paste0(year,"_",month,"_",day,"_", decimalLatitude,"_", decimalLongitude,"_", species))

@@

@@

### Comparison of drought frequency between annual and perennial species
```{r speciesmeans}
species_means <- GBIF_cleaned %>%
  dplyr::select(species, life_history, Winter, Summer, Spring, Fall) %>%
  group_by(species, life_history) %>%
  summarise(n=n(), Winter_mean=mean(Winter), Spring_mean=mean(Spring), Summer_mean=mean(Summer), Fall_mean=mean(Fall)) %>% 
  as.data.frame()

@@

species_means$life_history_num<-as.numeric(species_means$life_history)-1
row.names(species_means)<-gsub(" ", "_", species_means$species)
seasons<-c("Winter_mean", "Spring_mean", "Summer_mean","Fall_mean")

@@

```
To evaluate the hypothesis that annual and perennial life history strategies reflect adaptations to alternative drought regimens, we tested the corresponding prediction that the observed distributions of annual and perennial _Heliophila_ species would be significantly associated with historic drought frequency. We  tested for a relationship between drought frequency and life history, season, and their interaction by analysis of variance while including species as a random effect using the lme4 package in R [@bates2014fitting] and compared annuals and perennials using by Tukey adjusted post-hoc contrasts. We next calculated the mean drought frequency during the winter, spring, summer and fall for each species.  Because shared evolutionary history of closely related species can lead to spurious associations between traits and environments  [@felsenstein1985phylogenies], we tested for a relationship between life history strategy and drought frequency while controlling for phylogeny using phylogenetic logistic regression [@ives2010phylogenetic]. This statistical approach is designed to control for the confounding effects of common ancestry's influence on demographic features such as geospatial relationships when addressing hypotheses about the role of natural selection on trait distributions.

@@

# Results
(ref:phylogeny) Species and examples of herbaria specimens of _Heliophila_ (a) Phylogeny and life history strategies of species studied. Orange circles at branch tips mark annual species and blue circles mark perennial species. At internal nodes, pie charts indicate the estimated posterior probability of being annual versus perennial. Example herbaria specimens accessed via GBIF of (a) _H. minima_, (b) _H. deserticola_, (c) _H. coronopifolia_ and (d) _H. ephemera_. Images (b,d,e) courtesy of The Bavarian Natural History Collections (CC BY-SA 4.0) and (c) The London  Natural History Museum (CC BY 4.0). Links to images are found in the supplement.

@@

(ref:mapsboxplots) Locations of occurrence records of (a) annual and (b) perennial _Heliophila._ Drought frequency during the (c) winter, (d) spring, (e) summer and (f) fall detected using the VHI. (g) Drought frequencies during each season at the observation locations of annual and perennial _Heliophila_ (Post-hoc contrasts annuals and perennials from ANOVA, \*\* = p < 0.01).

@@

Fall_SA <-crop(Fall, rbind(c(12,37),c(-35, -18)))

fallmap<-gplot(Fall_SA) + 
  geom_tile(aes(fill=value)) + 
  scale_fill_gradientn(colors=col_scale, na.value="white", name=NULL) +
  geom_polygon(data = world, aes(x=long, y = lat, group=group),  fill = NA, color = "black") + 
  coord_fixed(xlim=c(13,35), ylim=c(-35, -22)) + 
  theme_classic()+
  theme(legend.position = c(0.95,.3),legend.text=element_text(size=rel(0.7)), legend.key.size=unit(.4,"line"), legend.background = element_rect(fill=alpha('white', 0)), text = element_text(size=8))+
  labs(title="Fall drought freq.", x="Longitude (°)", y="Latitude (°)")

GBIF_cleaned_long<- GBIF_cleaned %>%
  dplyr::select(species, life_history, Winter, Spring, Summer,Fall) %>%
  gather(season, drought_freq, Winter, Spring, Summer,Fall)

GBIF_cleaned_long$season<-factor(GBIF_cleaned_long$season, levels=c("Winter","Spring","Summer","Fall"))

ann_text <- data.frame(x = 1.5,drought_freq = 0.8,life_history="a", lab = "**",
                       season = factor(c("Summer","Fall"), levels=c("Winter","Spring","Summer","Fall")))

@@

Out of `r nrow(GBIF)` _Heliophila_ GBIF records,  `r sum(GBIF$species %in% species_life_history$species)` were for species with reported life history [@mummenhoff2005phylogeny], `r sum((GBIF$species %in% species_life_history$species) & !is.na(GBIF$decimalLatitude))` had geospatial data, `r sum((GBIF$species %in% species_life_history$species) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & !(is.na(GBIF$decimalLatitude)))` did not have geospatial issues, `r sum((GBIF$species %in% species_life_history$species) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & !(is.na(GBIF$decimalLatitude)) & !(is.na(GBIF$Summer)) & !(is.na(GBIF$decimalLatitude)))` were located on pixels classified as land having drought measurements, `r sum(GBIF$decimalLatitude < (-20) & GBIF$decimalLongitude < (40) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & (GBIF$species %in% species_life_history$species) & !(is.na(GBIF$Summer)))` were located in Africa, `r sum(!duplicated(GBIF$ymd_lat_lon_sp) & GBIF$decimalLatitude < (-20) & GBIF$decimalLongitude < (40) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & (GBIF$species %in% species_life_history$species) & !(is.na(GBIF$Summer))) ` were not duplicated. Thus, after all filtering steps, `r nrow(GBIF_cleaned)` records for `r length(unique(GBIF_cleaned$species))` species (Figure \@ref(fig:phylogeny), Table \@ref(tab:speciesmeanstable)) passed for further analyses. The number of samples varied between species, with a mean of `r mean(species_means$n)` samples per species.  _H. rigidiuscula_ had the most records, `r max(species_means$n)`,  and  _H. cornellsbergia_ the fewest,  `r min(species_means$n)` (Table \@ref(tab:speciesmeanstable)).

There were clear visual differences between the distributions of the `r sum(GBIF_cleaned$life_history=="a")` annual and the `r sum(GBIF_cleaned$life_history=="p")` perennial _Heliophila_ observation records (see Figure \@ref(fig:speciesmaps) for maps of individual species). While annual species were generally found in the western regions of South Africa and Namibia, primarily in the Cape Floristic Region and Succulent Karoo (Figure \@ref(fig:mapsboxplots)a), the occurrence of perennials extended to the southern and eastern coast of South Africa (Figure \@ref(fig:mapsboxplots)b).

@@

melted<-GBIF_cleaned %>% select(life_history, Winter, Spring, Summer,Fall, species) %>% gather(Winter, Spring, Summer,Fall, key="season",value="droughtfreq")

@@

post_hoc<-emmeans(fit, list(pairwise ~ life_history*season), adjust = "tukey")
post_hoc<-as.data.frame(post_hoc$`pairwise differences of life_history, season`)
post_hoc<-post_hoc[c(1,14,23,28), ]
post_hoc$contrast<-c("Fall","Spring","Summer","Winter")

@@

We found that the frequency of drought was significantly higher at the locations of occurrence records for annual species. When comparing across all occurrence records (all records rather than species means, Figure \@ref(fig:mapsboxplots)g), a mixed-model analysis of variance which included species as random effect revealed a significant relationship between drought frequency and life history (p < 0.01), season (p < 0.01) and their interaction (p < 0.01) (Table \@ref(tab:anovatable)). Post-hoc contrasts showed that the frequency of drought was significantly higher at the location of annuals during the summer (z ratio = `r post_hoc$z.ratio[post_hoc$contrast=="Summer"]`, p = < 0.01), and fall (z ratio = `r post_hoc$z.ratio[post_hoc$contrast=="Fall"]`, p < 0.01).  Because a comparison across all occurrence records does not account for variation in the number of records per species (Table \@ref(tab:speciesmeanstable)) or species relatedness (Figure \@ref(fig:phylogeny)a), we also tested whether mean drought frequency values of each species were significantly different between annuals and perennials using phylogenetic logistic regression. We found that the mean drought frequencies were significantly higher ($\alpha = 0.05$) in annual species during the spring, summer, and fall (Table \@ref(tab:modelstable), Figure \@ref(fig:lineplots)a,b). These findings indicate that common ancestry alone does not explain differences in the drought frequencies experienced between the environments of annual and perennial _Heliophila_.

@@

apa_table(modelstable
  , caption = "Phylogenetic logistic regressions between life history, and the mean drought frequency observed at specimen sites of Heliophila species the winter, spring, summer, and fall."
  , note = "Annual species were scored as 0 and perennial species as 1.",
  midrules=c(2,4,6,8),
  digits=4
)

@@

(ref:lineplots) (a) Comparison (mean +- SE) of drought frequency across seasons calculated at the occurrence locations of GBIF records of annual and perennial species of _Heliophila_. (b) Results from phylogenetic logistic regression, where the line shows the model fit and \* = p < 0.05, \*\* = p < 0.01. Annuals were scored as 0 and perennials as 1.  (c) Collection dates of GBIF records of annual and perennial species of _Heliophila_ in relation to the photoperiodic calendar where day 1 is intermediate to the fall equinox and winter solstice.

@@

life_history_drought_means<- species_means %>% 
  dplyr::select(life_history, Winter_mean, Spring_mean, Summer_mean,Fall_mean)  %>%
  gather(variable, value, -life_history) %>%
  group_by(life_history, variable) %>%
  summarise(mean = mean(value), se = sd(value)/sqrt(length(value))) %>%
  mutate(variable = reorder(variable, c(4,2,3,1)))

>>>>

author: 
  - name          : "J. Grey Monroe"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "307 University Ave, Fort Collins, CO 80521"
    email         : "monroejg@colostate.edu"
    phone         : "919-810-8800"
    orcid         : 0000-0002-4025-5572
    twitter       : "@Grey_Monroe"
  - name          : "Brian Gill"
    affiliation   : "3"
    twitter       : "@BrianGillPhD"
    orcid         : 0000-0001-9567-8989
  - name          : "Kathryn G. Turner"
    affiliation   : "4"
    twitter       : "@KTInvasion"
    orcid         : 0000-0001-8982-0301
  - name          : "John K. McKay"
    affiliation   : "2"
    orcid         : 0000-0003-4311-5513

@@

keywords          : "drought adaptation, herbaria records, Heliophila, life history evolution, phylogeography, remote sensing"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word

@@

Understanding the causes of life history variation and climate adaptation are longstanding goals of ecology and evolutionary biology  [@raunkiaer1905types;@turesson1925plant;@clausen1948experimental;@cole1954population]. In  plants, life histories are especially diverse, with some species completing their life cycle in a number of weeks to others that live for thousands of years [@brown1996oldlist]. Along this continuum in angiosperms an important division exists distinguishing annuals which complete their seed to seed life cycle within a single year from perennials which can persist over multiple years. Annual plants do not need to survive through the full range of seasonal environmental variation and spend at least some portion of the year as a seed where they are relatively protected from environmental stress. In contrast, perennial plants can continue vegetative growth over multiple years and must survive conditions experienced during all seasons but can also benefit from competitive advantages and, if iteroparous, multiple bouts of reproduction. These represent fundamentally different life history strategies and predicting their occurrence is important for community, ecosystem, and agricultural ecology. However, the environmental factors that explain their evolution and distributions remain empirically unresolved [@friedman2015all].

Classical theory predicts shorter life spans in environments where adult mortality is high [@charnov1973life;@stearns1992evolution;@franco1996life]. Because lack of water is one of the greatest threats to survival during vegetative or reproductive growth in plants, this theory has been extended to the hypothesis that annuality is adaptive when it allows plants to escape drought [@cody1975ecology;@schaffer1975selection]. Indeed, adaptation to drought,  defined as episodes of increased aridity causing plant stress [@passioura1996drought], is often invoked as an explanation for the success of annual species, though drought adapted perennials are also well known [@raunkiaer1905types]. And while a few cases are cited where annuality appears to be more common in environments with greater aridity [@stebbins1952aridity; @morishima1984differentiation], this hypothesis has yet to be supported while controlling for the effect of common ancestry (phylogeny) on life habit. In one previous study where this question was addressed phylogenetically,  [@evans2005climate] annuals were not found to be associated with environments that experience more drought. This could be explained by the relatively small number of species studied and the reliance on a limited number of weather stations to characterize environments, highlighting the need to develop more scalable methods to study the geographic distributions of traits such as life history. Thus, in this study we leverage thousands of herbaria specimens among dozens of species and high-resolution remote sensing to study the distributions and environmental factors potentially driving the evolution and distribution of life history.

It is also critical to consider another dimension of drought adaptation: the expectation that annuality is most adaptive when droughts are not only frequent but also predictable. That is, when the frequency of drought is particularly high during certain seasons. Such predictability is important for selection to favor an escape strategy during those seasons which are particularly drought prone. While there has been at least one example of annuality associated with environments qualitatively classified as "predictable" in a general sense [@datson2008climate], the seasonal predictability of drought experienced by annuals has yet to be rigorously studied. As such, further empirical work is needed to support the model of annuality as a mechanism of drought adaptation via escape from drought prone seasons. Here we examine herbarium collection dates to ask whether annuals indeed exhibit evidence of an escape strategy from seasons with elevated drought frequency.

@@

Here we combine a long-term global dataset of satellite detected drought events with metadata from natural history collections to test these classic hypotheses within the African endemic mustard genus, _Heliophila_ L. (Brassicaceae). If annuality is an adaptive strategy allowing plants to escape drought prone seasons, then drought frequency should predict the distribution of life history strategies across landscapes, and annual species should be more commonly associated with drought prone regions than perennial species. Additionally, if perenniality offers competitive advantage in the absence of drought, associations between life history and drought frequency should be significant when phylogenies include transitions from annual to perennial life history strategy. Finally, if annual species have adapted to escape predictably drought prone seasons, observations of growing annual species (i.e. occurring in forms other than seed) should be rare during seasons when drought frequency is highest. Phylogenetic relatedness can influence tests of associations between species’ traits and their environments by confounding common environments caused by selection from common environments caused by ancestry. [@barrett1996comparative;@felsenstein1985phylogenies]. Therefore, we assessed the relationship between life history distribution and drought frequency while controlling for phylogeny.

@@

### _Heliophila_ occurrence records
```{r raw_GBIF}
GBIF<-read.csv("../data/0006178-160311141623029/occurrence.csv")
```
To characterize the distributions of annual (studied here, n = 21) and perennial (studied here, n = 21) _Heliophila_ species, all (`r nrow(GBIF)`) records for the genus _Heliophila_ were downloaded from the Global Biodiversity Information Facility (gbif.org) on July 21, 2018 [@gbifdownload]. Herbaria records such as these provide a rich data sources to characterize the geographical distributions of species [@thiers2016index; @willis2017old; @lang2019using]. And as they become digitized [@soltis2017digitization], herbaria collections have been used to study relationships between trait distributions, geography, and climate [@davis2015herbarium;@wolf2016altitudinal;@stropp2016mapping;@vaclavik2017effects].

@@

```{r Droughtdata}
Winter<-raster("../data/drought/Summer_drought_freq")
Spring<-raster("../data/drought/Fall_drought_freq")
Summer<-raster("../data/drought/Winter_drought_freq")
Autumn<-raster("../data/drought/Spring_drought_freq")

@@

$$VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})$$
By combining VCI and TCI, the VHI distinguishes drought from other forms of vegetative stress [@kogan1995application]. The use of the VHI to detect drought has been validated globally and across ecosystem types [@aghakouchak2015remote], including in southern Africa, the focal region of this study (e.g. Figure \@ref(fig:mapsdroughtexamples)). To date, the VHI has most often been applied for evaluating drought risk for agricultural research [e.g., @rojas2011assessing; @kogan2016modelling]. But it also presents a tool to study seasonal patterns in the frequency of drought across environments and to test hypotheses about the effect of drought on ecological and evolutionary processes [@kerr2003space]. As such, the VHI  has been applied recently to study drought related ecology of natural species and proven useful for predicting intraspecific variation in drought tolerance traits and genes [@mojica2016genetics;@dittberner2018natural;@monroe2018drought]. Here, we accessed VHI data at $16km^2$ resolution from 1981 to 2015 (https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php) to characterize the seasonal drought frequencies experienced by annual and perennial _Heliophila_ species across their native range of southern Africa.

## Analyses
### Drought frequency calculations
To characterize drought regimens across the distributions of annual and perennial species of _Heliophila_, we calculated drought during different seasons at the location of observations for _Heliophila_ records using the VHI. Specifically, we created maps of the frequencies of observing drought conditions between years (VHI<40, NOAA) during the winter (quarter surrounding winter solstice), spring (quarter surrounding spring equinox), summer (quarter surrounding summer solstice) and autumn (quarter surrounding autumn equinox) from 1981 to 2015 across the range of _Heliophila_. From these maps, the drought frequency (the number of times drought is observed divided by the total number of years, 34) during the winter, spring, summer, and autumn were extracted for the locations of all GBIF records.

@@

GBIF <- GBIF %>%
  mutate(Winter=raster::extract(Winter, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(Spring=raster::extract(Spring, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(Summer=raster::extract(Summer, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(Autumn=raster::extract(Autumn, cbind(decimalLongitude, decimalLatitude))) %>%
  mutate(ymd_lat_lon_sp = paste0(year,"_",month,"_",day,"_", decimalLatitude,"_", decimalLongitude,"_", species))

@@

Out of `r nrow(GBIF)` _Heliophila_ GBIF records,  `r sum(GBIF$species %in% species_life_history$species)` were for species with reported life history [@mummenhoff2005phylogeny], `r sum((GBIF$species %in% species_life_history$species) & !is.na(GBIF$decimalLatitude))` had geospatial data, `r sum((GBIF$species %in% species_life_history$species) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & !(is.na(GBIF$decimalLatitude)))` did not have geospatial issues, `r sum((GBIF$species %in% species_life_history$species) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & !(is.na(GBIF$decimalLatitude)) & !(is.na(GBIF$Summer)) & !(is.na(GBIF$decimalLatitude)))` were located on pixels classified as land having drought measurements, `r sum(GBIF$decimalLatitude < (-20) & GBIF$decimalLongitude < (40) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & (GBIF$species %in% species_life_history$species) & !(is.na(GBIF$Summer)))` were located in Africa, `r sum(!duplicated(GBIF$ymd_lat_lon_sp) & GBIF$decimalLatitude < (-20) & GBIF$decimalLongitude < (40) & !is.na(GBIF$decimalLatitude) & !(GBIF$hasGeospatialIssues==T) & (GBIF$species %in% species_life_history$species) & !(is.na(GBIF$Summer))) ` were not duplicated.

@@

### Comparison of drought frequency between annual and perennial species
```{r speciesmeans}
species_means <- GBIF_cleaned %>%
  dplyr::select(species, life_history, Winter, Summer, Spring, Autumn) %>%
  group_by(species, life_history) %>%
  summarise(n=n(), Winter_mean=mean(Winter), Spring_mean=mean(Spring), Summer_mean=mean(Summer), Autumn_mean=mean(Autumn)) %>% 
  as.data.frame()

@@

species_means$life_history_num<-as.numeric(species_means$life_history)-1
row.names(species_means)<-gsub(" ", "_", species_means$species)
seasons<-c("Winter_mean", "Spring_mean", "Summer_mean","Autumn_mean")

@@

```
To evaluate the hypothesis that annual and perennial life history strategies reflect adaptations to alternative drought regimens, we tested the corresponding prediction that the observed distributions of annual and perennial _Heliophila_ species would be significantly associated with historic drought frequency. We  tested for a relationship between drought frequency and life history, season, and their interaction by analysis of variance while including species as a random effect using the lme4 package in R [@bates2014fitting] and compared annuals and perennials using Tukey adjusted post-hoc contrasts. We next calculated the mean drought frequency during the winter, spring, summer and autumn for each species.  Because shared evolutionary history of closely related species can lead to spurious associations between traits and environments  [@felsenstein1985phylogenies], we tested for a relationship between life history strategy and drought frequency while controlling for phylogeny using phylogenetic logistic regression [@ives2010phylogenetic]. This statistical approach is designed to control for the confounding effects of common ancestry's influence on demographic features such as geospatial relationships when addressing hypotheses about the role of natural selection on trait distributions.

@@

# Results
(ref:phylogeny) Species and examples of herbaria specimens of _Heliophila_ (a) Phylogeny and life history strategies of species studied. Orange circles at branch tips mark annual species and blue circles mark perennial species. At internal nodes, pie charts indicate the estimated posterior probability of being annual versus perennial. Example herbaria specimens accessed via GBIF of (b) _H. minima_, (c) _H. deserticola_, (d) _H. coronopifolia_ and (e) _H. ephemera_. Images (b,d,e) courtesy of The Bavarian Natural History Collections (CC BY-SA 4.0) and (c) The London  Natural History Museum (CC BY 4.0). Links to images are found in the supplement.

@@

(ref:mapsboxplots) Locations of occurrence records of (a) annual and (b) perennial _Heliophila._ Drought frequency during the (c) winter, (d) spring, (e) summer and (f) autumn detected using the VHI. (g) Drought frequencies during each season at the observation locations of annual and perennial _Heliophila_ (Post-hoc contrasts annuals and perennials from ANOVA, \*\* = p < 0.01). Boxplots indicate median (horizontal line), first and third quartiles (upper and lower limits of box), and 1.5 times the interquartile range (whiskers). Individual observations are shown by crosshairs. Annual and perennial species are indicated by data colored blue and oranage, respectively.

@@

Fall_SA <-crop(Autumn, rbind(c(12,37),c(-35, -18)))

fallmap<-gplot(Fall_SA) + 
  geom_tile(aes(fill=value)) + 
  scale_fill_gradientn(colors=col_scale, na.value="white", name=NULL) +
  geom_polygon(data = world, aes(x=long, y = lat, group=group),  fill = NA, color = "black") + 
  coord_fixed(xlim=c(13,35), ylim=c(-35, -22)) + 
  theme_classic()+
  theme(legend.position = c(0.95,.3),legend.text=element_text(size=rel(0.7)), legend.key.size=unit(.4,"line"), legend.background = element_rect(fill=alpha('white', 0)), text = element_text(size=8))+
  labs(title="Autumn drought freq.", x="Longitude (°)", y="Latitude (°)")

GBIF_cleaned_long<- GBIF_cleaned %>%
  dplyr::select(species, life_history, Winter, Spring, Summer,Autumn) %>%
  gather(season, drought_freq, Winter, Spring, Summer,Autumn)

GBIF_cleaned_long$season<-factor(GBIF_cleaned_long$season, levels=c("Winter","Spring","Summer","Autumn"))

ann_text <- data.frame(x = 1.5,drought_freq = 0.8,life_history="a", lab = "**",
                       season = factor(c("Summer","Autumn"), levels=c("Winter","Spring","Summer","Autumn")))

@@

After all filtering steps, `r nrow(GBIF_cleaned)` records for `r length(unique(GBIF_cleaned$species))` species (Figure \@ref(fig:phylogeny), Table \@ref(tab:speciesmeanstable)) passed for further analyses. The number of samples varied between species, with a mean of `r mean(species_means$n)` samples per species.  _H. rigidiuscula_ had the most records, `r max(species_means$n)`,  and  _H. cornellsbergia_ the fewest,  `r min(species_means$n)` (Table \@ref(tab:speciesmeanstable)).

There were clear visual differences between the distributions of the `r sum(GBIF_cleaned$life_history=="a")` annual and the `r sum(GBIF_cleaned$life_history=="p")` perennial _Heliophila_ observation records (Figure \@ref(fig:mapsboxplots), Figure \@ref(fig:speciesmaps)). While annual species were generally found in the western regions of South Africa and Namibia, primarily in the Cape Floristic Region and Succulent Karoo (Figure \@ref(fig:mapsboxplots)a), the occurrence of perennials extended to the southern and eastern coast of South Africa (Figure \@ref(fig:mapsboxplots)b).

@@

melted<-GBIF_cleaned %>% select(life_history, Winter, Spring, Summer,Autumn, species) %>% gather(Winter, Spring, Summer,Autumn, key="season",value="droughtfreq")

@@

post_hoc<-emmeans(fit, list(pairwise ~ life_history*season), adjust = "tukey")
post_hoc<-as.data.frame(post_hoc$`pairwise differences of life_history, season`)
post_hoc<-post_hoc[c(1,14,23,28), ]
post_hoc$contrast<-c("Autumn","Spring","Summer","Winter")

@@

We found that the frequency of drought was significantly higher at the locations of occurrence records for annual species. When comparing across all occurrence records (all records rather than species means, Figure \@ref(fig:mapsboxplots)g), a mixed-model analysis of variance which included species as random effect revealed a significant relationship between drought frequency and life history (p < 0.01), season (p < 0.01) and their interaction (p < 0.01) (Table \@ref(tab:anovatable)). Post-hoc contrasts showed that the frequency of drought was significantly higher at the location of annuals during the summer (z ratio = `r post_hoc$z.ratio[post_hoc$contrast=="Summer"]`, p = < 0.01), and autumn (z ratio = `r post_hoc$z.ratio[post_hoc$contrast=="Autumn"]`, p < 0.01).  Because a comparison across all occurrence records does not account for variation in the number of records per species (Table \@ref(tab:speciesmeanstable)) or species relatedness (Figure \@ref(fig:phylogeny)a), we also tested whether mean drought frequency values of each species were significantly different between annuals and perennials using phylogenetic logistic regression. We found that the mean drought frequencies were significantly higher ($\alpha = 0.05$) in annual species during the spring, summer, and autumn (Table \@ref(tab:modelstable), Figure \@ref(fig:lineplots)a,b). These findings indicate that common ancestry alone does not explain differences in the drought frequencies experienced between the environments of annual and perennial _Heliophila_.

@@

apa_table(modelstable
  , caption = "Phylogenetic logistic regressions between life history, and the mean drought frequency observed at specimen sites of Heliophila species the winter, spring, summer, and autumn"
  , note = "Annual species were scored as 0 and perennial species as 1.",
  midrules=c(2,4,6,8),
  digits=4
)

@@

(ref:lineplots) (a) Comparison (mean +- SE) of drought frequency across seasons calculated at the occurrence locations of GBIF records of annual and perennial species of _Heliophila_. (b) Results from phylogenetic logistic regression, where the line shows the model fit and \* = p < 0.05, \*\* = p < 0.01. Annuals were scored as 0 and perennials as 1.  (c) Collection dates of GBIF records of annual and perennial species of _Heliophila_ in relation to the photoperiodic calendar where day 1 is intermediate to the autumn equinox and winter solstice.

@@

life_history_drought_means<- species_means %>% 
  dplyr::select(life_history, Winter_mean, Spring_mean, Summer_mean,Autumn_mean)  %>%
  gather(variable, value, -life_history) %>%
  group_by(life_history, variable) %>%
  summarise(mean = mean(value), se = sd(value)/sqrt(length(value))) %>%
  mutate(variable = reorder(variable, c(4,2,3,1)))
